---
title: "Introduction to R"
author: "Brian Capaldo"
date: "Saturday June 10, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data structures

R offers a variety of built-in data structures.

### Base types

The base types include `double`, `integer`, `characater` and `logical`.  Values can be tested as belonging to one of those types.

```{r, eval=F}
typeof(1)
## [1] "double"
typeof(1L)
## [1] "integer"
is.numeric(1.1)
## [1] TRUE
is.logical(TRUE)
is.character("Hello")
```
Every base type has a sentinel value used to represent missing observations.

```{r,eval=F}
x <- 1
x[2] <- NA
typeof(x[2])
## "double"
typeof(NA)
## logical
typeof(NA + 1L)
```

### Vectors

R provides several kinds of compound data structures commonly referred to as *vectors*.  All operations in R are *vectorized*, this means that they take vectors as arguments and they will operate on the individual values (basic types) of these vectors.

```{r}
x <- c(1,2,2)
y <- c(2,2,1)
( x + y ) * x
```

All vectors can be accessed used subsetting operators `[` and `[[`, and can be used with functions `typeof()` (to find out what basic type is in the vector),
`length()` (how many elements are in it), `attributes()` (query the vector's meta data).

#### Atomic vectors

Atomic vectors are homogeneous (possibly multi-dimensional) data structures.

```{r}
c(1,2,3)
c("one","two","three")
```

Element of vectors can be accessed individually.

```{r}
x <- c(1L, 2L, 3L)
x[1] <- 45
x[1]
```

The operator `[<-` is the assignment function.

A two dimensional atomic vector is called a *matrix* and higher-dimensional vector is called an *array*.

```{r}
a <- matrix(1:6, ncol = 3, nrow = 2)
b <- array(1:12, c(2,3,4))
a
b
```

Matrices and array, like vectors, are homogeneous, but it is possible to assign values of different basic types into any one of them. This causes a *coercion* of the entire data structure.

```{r}
x[[1]] <-1.1
x
a[[2,1]] <- "one"
a
```

Coercions go from least to most flexible:  logical, integer, double and character.

**Exercise 8:** what should `a[2,1] <- 1` do?


R has a powerful set of subsetting operations that apply to all vectors uniformly. R allows, to subset ranges, rows, columns of vectors.

```{r}
a[1,]
a[1:2,]
a[1:2,2:3]
aa <- as.integer(a)
dim(aa) <- c(2,3)
a[1, a[1,] > 1]
aa[1, aa[1,] > 1]
```

When using a predicate such as `a[1,] > 1` for subsetting, behind the scenes R generates a vector of logical values. True indicates a position that should be extracted.  So this is:

```{r}
aa[1,]>1
```

We can use logical arrays to subset as follows:

```{r}
aa[1, c(FALSE, TRUE, TRUE)]
```

Subsetting can be used in assignment operations as well.

```{r}
a[1:3] <- a[4:6]
```


#### Lists

Lists are heterogeneous vectors. The elements of list can be of any kind, including lists and vectors.

```{r}
list(1, "hi", c(1,2))
```

The `typeof()` a list is `list`. You can test for a list with `is.list()` and coerce to a list with `as.list()`. You can turn a list into an atomic vector with `unlist()`. If the elements of a list have different types, `unlist()` will coerce.

### Referential transparency

To facilitate equational reasoning, R attempts to provide *referential transparency* for function calls. Referential transparency means that arguments are not changed by the function being called. So the following function `f(x)` does not modify the vector passed in.

```{r}
x
f <- function(x) { x[1] <- 1 }
f(x)
x
```

But then what does it change in the body?

The `pryr` package provides function `object_size` to estimate the size of several data structures.

```{r}
X <- 1:(10*1000*1000)
if (!require("pryr",character.only = TRUE)) install.packages("pryr")
pryr::object_size(X)
Y <- X
pryr::object_size(X,Y)
Y[1] <- 11
pryr::object_size(X,Y)
```

**Exercise 9:** Explain the two surprising things about the above results.

## Attributes

Any vector in R can have attributes. Attributes are a named `list`. Attributes can be accessed individually with `attr()` or all at once (as a list) with `attributes()`.

```{r}
y <- 1:10
attr(y, "my_att") <- "This is a vector"
```

Attributes can be set and queried.

```{r}
attr(y, "my_att")
```

The `str()` function clearly shows that attributes are implemented by a `list`.

```{r}
str(attributes(y))
```

Some attributes such as `dim`, `name`, `class` are understood by many R functions.  For instance, setting the `names` attribute allows to name positions in a vector (and row/column in matrices).

```{r}
x <- c(1,2,3)
attr(x,"names") <- c("a","b","c")
x["a"]
```

## Delayed Evaluation

**Exercise 10:** Consider the following code, what will it print?

```{r}
print(1)   # outputs "1" on the terminal
f<-function(a,b) { invisible(0) }  # a helper, nevermind
f( print(1) , print(2) )   # what would this do in Java or in C
temp <-print(1);temp2 <- print(2); f(temp, temp2)  # same thing?
```

**Exercise 11:** What will this print?

```{r}
f1 <- function( a, b, c ) { # Consider the following, not very useful function
 x <- f(b,b)
 x <- c
 x <- c
 a
}

f1( print(1), print(2), print(3) ) # What do you recon this will print?
```

**Exercise 12:** What will this print?

```{r}
f2 <- function( a, b, c ) { # here is another function
  x <- b
  x <- c
  f3(a)
}

f3 <- function( a) { } # with a silly helper function

f2( print(1), print(2), print(3) ) # what will we print?
```

In R, arguments to a functions are not evaluated, they are packaged into *promises*. A promise is only evaluated if it is accessed (and only the first time it is accessed).  

Every expression is passed as a promise, except constants.

The following is an obscure corner of R.   Enjoy.

```{r}
FU <- function(b="") print(paste("FU",b))                # A function
BAR <- function( FU = print("BAR"))  FU("bar")   # Another function with a default argument
BAR()   # call bar will print ?
BAR( function(n) print("YO") )  # and now?
```


# Building Abstractions

We have now reviewed the basic building blocks of the language. Using these building blocks it is possible to construct powerful abstractions.

## Building a factor abstraction

First, lets look at how to implement factors using attributes.
 
```{r}
simple_factor <- function (x) {
    levels <- as.character(unique(x))
    f <- match(x, levels)
    attr(f, "levels") <- levels
    attr(f, "class") <- "simple_factor"
    f
}

x <- c(42, 42, 2, 3, 4)
simple_factor(x)
```
Printing is not very elegant. Let's add a dedicated print function.

```{r}
print.simple_factor <-  function (x, ...)  {
   xx <- x  # copy x
   class(xx) <- NULL  # delete its class
   levels(xx) <- NULL # delete its levels
   xx <- sapply(xx, function(y) attr(x,"levels")[y]) # replace levels with their values
   print(xx, quote=FALSE)   # print without quotes
   cat(paste("Levels: ", paste(levels(x),collapse=" ")))  #print the levels
  invisible(x)  # return x without echoing it
}

print(simple_factor(x))
```

## Building a counter abstraction

With referential transparency it is difficult to mutate non-local state.  What if we wanted to create a large number of counters and did not want anyone but us to be able to update them.



```{r}
remove(xPriv)
counter <- function() { xPriv <- xPriv + 1; xPriv}
#counter()             # oops, we need to define x
xPriv<-0
counter()
counter()             # Hmm... not quite

# Super assignment to the rescue
counter <- function() { xPriv <<- xPriv + 1; xPriv}  # <<- assigns in the parent env
counter()
counter()             # now we have a counter
```

This only lets us have a single counter. What if we want a counter factory?


```{r}
mkC <- function () { 
    xPriv <- 0
    function() { xPriv <<- xPriv + 1; xPriv }
  }

count <- mkC()
count()
count()
```

The above uses the environment of the `mkC()` function to store the counter. Each time we call `mkC`, a new environment is created.


## Building abstractions with Objects

R has multiple object systems. They are constructed on top of attributes and reflective method invocation.

The S3 system gives us single dispatch and simple structures.


```{r}
x<- 1
attr(x, "class" ) <- "pen"      # class is just an attribute like any other
x
y<- 1
class(y ) <- "cowboy"           # there is a dedicate function to set
y
class(y) <- c("cow?cboy","human") # one can be several things

draw <- function( this )  UseMethod("draw", this)  # this introduces the generic method
draw.cowboy <- function(this)  "Bang"      # different variants for different classes
draw.pen <- function(this) "Scribble"
draw.default <- function(this) "Huh?"     # default version when nothing else applies

draw(x)
draw(y)
draw(1)
```

The S4 system is much richer.

```{r}
setClass("Person",   # Create an new class with two slots
      slots = list(name = "character", age = "numeric"))

setClass("Employee",  # Extends the Person class with another class
      slots = list(boss = "Person"), 
      contains = "Person") # declare an inheritance relationship

# create a few people
alice <- new("Person", name = "Alice", age = 40)
john <- new("Employee", name = "John", age = 20, boss = alice) # john's got a boss
jack <- new("Person", name = "Jack", age = 42)

alice@age              # ask an impolite question
slot(john, "boss")     # same

# Let's define methods
setGeneric("hire",  function(x,y) standardGeneric("hire") )
# The first will work only for persons hiring employees
setMethod("hire", 
       c(x = "Person", y = "Employee"), 
       function(x, y) {
         print(paste(x@name," hired ",y@name)) 
         return(setBoss(y)<-x)  # change the boss slot
       }
)

# another method for setting fields
setGeneric("setBoss<-",function(object,value){standardGeneric("setBoss<-")})
setReplaceMethod(f="setBoss",
              signature="Employee",
              definition=function(object,value) return(object@boss <-value)
              )

hire(alice, john)   # this should work

print(hire(jack, john))  # and this too, changing the boss

hire(john, john)  # this should work as an employee is person
#hire(alice, alice) # this should fails as an employee isn't a person
```

## Build abstraction with domain specific languages

R allows to leverage delayed evaluation in order to implement domain specific languages.

Here is the example of timer function that repeatedly run the expression passed in as argument and reports the timing of that expression.

```{r}
tm <- function( exp ) {
 x<- deparse(substitute( exp ))   ## decompile the expression
 t <- 1                           ## repetion counter (we run exp 10 * t)
 time = 0;                        ## reset timer
 while ( time < 1 ) {             ## continue looking until the code runs long enough
  y <- x                          ## code
  for( i in 1:9 ) y <- paste(y,";",x)  ## duplicate the code 10 times
  y<- paste(" system.time ( for (i in 1:",t,") {", 
            y,                    ## wrap the code with a loop that runs t times
            "}, gcFirst= FALSE)")
  r <- eval(parse(text=y))        ## execute
  time <- r[3]                    ## grab the elasped time
  t <- t * 10                     ## try with a bigger t
 }
 (time / (t * 10)) * 1000         ## return time in milliseconds (default is in s)
}

tm(1 + 1)
```
