---
title: "Introduction to R: A Computational Workbench for Biological Data Analysis"
author: "Brian Capaldo"
date: "6/10/2017"
output: slidy_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## History of R

- The R programming language has its roots in S, a language created forty years ago at ATT Bell Labs by Chambers, Becker and Wilks.

- The motivation for S was to simplify the task of calling statistical routines written in FORTRAN (this involved writing code to read in, format and write out the data).

- The goal of S was to free users of the drudgery of writing that interface code and speed up the data analyst’s workflow. The key linguistic design principle was to design a language that could be extended and that was devoid of arbitrary limitations.

- The S language was such a success that users started writing statistical routines in S.

- The R programming language was designed by Ross Ihaka and Robert Gentleman at the University of Auckland around 1992.

- R departed from S in number of ways, it was an open source language released under a GPL license to ensure that everything related to R remains in the public domain, R cleaned up some corners of S and improve the performance of programs written in the language.

## R as a computational bench top 

- Data as experiments
- Data structures as glassware
- Functions as instruments
- Vignettes as protocols

## Data types in R

The base types include `double`, `integer`, `characater` and `logical`.  Values can be tested as belonging to one of those types.

##

```{r, echo = TRUE}
typeof(1)
```

##

```{r, echo = TRUE}
typeof(1L)
```

##

```{r, echo = TRUE}
is.numeric(1.1)
```

##

```{r, echo = TRUE}
is.logical(TRUE)
is.character("Hello")
```


## Vectors

R provides several kinds of compound data structures commonly referred to as *vectors*.  All operations in R are *vectorized*, this means that they take vectors as arguments and they will operate on the individual values (basic types) of these vectors.

The operator `[<-` is the assignment function.

```{r, echo = TRUE}
x <- c(1,2,2)
y <- c(2,2,1)
( x + y ) * x
```

## Atomic vectors

Atomic vectors are homogeneous (possibly multi-dimensional) data structures.

```{r, echo = TRUE}
c(1,2,3)
c("one","two","three")
```

## Subsetting

All vectors can be accessed used subsetting operators `[` and `[[`, and can be used with functions `typeof()` (to find out what basic type is in the vector),
`length()` (how many elements are in it), `attributes()` (query the vector's meta data).

## Element of vectors can be accessed individually.

```{r, echo = TRUE}
x <- c(1L, 2L, 3L)
x[1] <- 45
x[1]
```

## A two dimensional atomic vector is called a *matrix* and higher-dimensional vector is called an *array*.

```{r, echo = TRUE}
a <- matrix(1:6, ncol = 3, nrow = 2)
b <- array(1:12, c(2,3,4))
```

## a

```{r, echo = TRUE}
a
```

## b

```{r, echo = TRUE}
b
```

## Dynamic typing and coercions 

Matrices and array, like vectors, are homogeneous, but it is possible to assign values of different basic types into any one of them. This causes a *coercion* of the entire data structure.

```{r, echo = TRUE}
x
x[[1]] <-1.1
x
```

##

```{r, echo=T}
a
a[[2,1]] <- "one"
a
```

## R has a powerful set of subsetting operations that apply to all vectors uniformly. R allows, to subset ranges, rows, columns of vectors.

```{r, echo = TRUE}
a[1,]
a[1:2,]
a[1:2,2:3]
```

## Coercions

```{r, echo = TRUE}
aa <- as.integer(a)
a
aa
```

##

```{r, echo = TRUE}
aa
dim(aa) <- c(2,3)
aa
a[1, a[1,] > 1]
aa[1, aa[1,] > 1]
```

## When using a predicate such as `a[1,] > 1` for subsetting, behind the scenes R generates a vector of logical values. True indicates a position that should be extracted.  So this is:

```{r, echo = TRUE}
aa[1,]>1
```

## We can use logical arrays to subset as follows:

```{r, echo = TRUE}
aa[1, c(FALSE, TRUE, TRUE)]
```

## Subsetting can be used in assignment operations as well.

```{r, echo = TRUE}
a
a[1:3] <- a[4:6]
a
```


## Lists

Lists are heterogeneous vectors. The elements of list can be of any kind, including lists and vectors.

```{r, echo = TRUE}
list(1, "hi", c(1,2))
```

The `typeof()` a list is `list`. You can test for a list with `is.list()` and coerce to a list with `as.list()`. You can turn a list into an atomic vector with `unlist()`. If the elements of a list have different types, `unlist()` will coerce.

## Referential transparency

To facilitate equational reasoning, R attempts to provide *referential transparency* for function calls. Referential transparency means that arguments are not changed by the function being called. So the following function `f(x)` does not modify the vector passed in.

```{r, echo = TRUE}
x
f <- function(x) { x[1] <- 1 }
f(x)
x
```

## Questions on the basics of R?

## Packages

- In R, collections of data, functions, and compiled code are stored as *packages*. *Packages* can be thought of as the computational benchtop equivalent of the reagents and instruments required to perfrom a specific computational assay. 

- There are two primary repositories from which to get *packages*, [CRAN](https://cran.r-project.org/) and [Bioconductor](http://bioconductor.org/)

- Bioconductor has made available a number of previous [courses](http://bioconductor.org/help/course-materials/) as well

## Installing packages from CRAN

```{r, echo=TRUE}
install.packages("tidyverse", verbose = T, repos='http://cran.us.r-project.org')
```

## Installing packages from Bioconductor

```{r, echo =T}
source("https://bioconductor.org/biocLite.R")
biocLite("flowCore", verbose = T, suppressUpdates = T)
```

## Loading libraries

```{r, echo = T}
library(flowCore)
library(tidyverse)
```

